---
title: "Week 5, Notes & Exercises"
subtitle: "Statistical Thinking (ETC2420 / ETC5242)"
date-format: "[Semester 2,] YYYY"
date: "2025"
toc: true
format:
  html:
    embed-resources: true
---

This week we apply the bootstrap method to calculate confidence intervals for
several different parameters. We do so using the CBT data from Week 3. Along
the way, we will practise using functions and loops in R.

```{r}
#| message: false
library(tidyverse)
set.seed(24205242)
```

# Visualising empirical distributions

In the exercises we do later below, there will be multiple times when we want
to visualise the distribution of a set of values. Rather than writing the same
code multiple times, we can write a function that does what we want and then
simply call that function whenever we need it.

The code chunk below defines a function for this purpose. However, some of the
code has been omitted. Can you fill in the missing parts?

***Exercise 1***

1.  Complete the code below to define a working function.

You'll need to change every instance of "`AAA`" with some suitable R code.

```{r}
#| eval: false
# Creates a histogram and smoothed histogram given a vector of values.
dataplot <- function(x, bins = 50, colour = "blue") {
  ggplot(tibble(x = x), AAA(x = x, y = after_stat(density))) +
    geom_AAA(bins = bins, colour = colour, fill = colour, alpha = 0.2) +
    geom_AAA(colour = colour, fill = colour, alpha = 0.2) +
    theme_bw()
}
```

Try out your function on some randomly generated data:

```{r}
#| eval: false
dataplot(rnorm(1000))
```

# Cognitive behaviour therapy

The following exercises will guide you through the process of applying the
bootstrap method to calculate a confidence interval.

Use the same `CBT.csv` data file that we used in the Week 3 tutorial. Some of
the exercises below repeat steps or ideas from Week 3.

## Difference in means

Initially, we will estimate the difference between two means,
$\delta = \mu_2 - \mu_1$, like we did in Week 3.

***Exercises 2--9***

2.  Read in the `CBT.csv` data file into a tibble called `CBT`.

3.  Create a new column called `Diff` which is the difference in scores for
    each subject, in the same way as in the Week 3 tutorial.

4.  Draw a histogram of these differences, overlayed with a smoothed histogram.

5.  Is it possible to also overlay the population distribution on this plot? If
    yes, then do so. If not, then explain why.

6.  Calculate the 95% confidence interval for $\delta$ using methods you learnt
    in Week 3. (This is for comparing against the bootstrap method, which we do
    below.)

7.  Starting with the `Diff` variable, generate $B = 5000$ bootstrap samples.
    For each bootstrap sample, calculate and save the sample mean, storing them
    in a vector (that vector should be of length 5000, with one value per
    bootstrap sample). You will need this vector for later exercises. Remember
    to set a seed so that your bootstrap samples are reproducible.

Some code to get you started is below, but you'll need to change every instance
of "`AAA`" with some suitable R code.

```{r}
#| eval: false
B <- AAA
xbar_boot <- rep(NA, B)

# Generate the bootstrap samples and calculate statistics.
for(b in 1:B) {
  AAA <- sample(AAA)
  xbar_boot[b] <- AAA
}
```

8.  Calculate a 95% confidence interval for $\delta$ from your bootstrap
    samples, using the percentile bootstrap method. (Hint: use the `quantile()`
    function.)

9.  The $B$ sample means of the bootstrap samples provide us with an empirical
    estimate of the sampling distribution of the `Diff` statistic. Let's
    visualise it. Draw a histogram of these $B$ values and overlay it with a
    smoothed histogram. Then also add the following elements:

    -   Vertical lines to show the lower and upper bounds of the 95% confidence
        intervals you calculated above. Use different colours and line styles
        to distinguish between the bootstrap-based CI (from Exercise 8) and the
        one you calculated earlier (from Exercise 6).
    -   An appropriate title and axis labels.

Some code to get you started is below, but you'll need to change every instance
of "`AAA`" with some suitable R code, and add any extra code as appropriate.

```{r}
#| eval: false
dataplot(xbar_boot, bins = 100) +
  geom_vline(AAA) +
  xlab(AAA) +
  ggtitle(AAA)
```

## Median difference

We now wish to estimate a different parameter: the median of the pairwise
difference in scores. Let $A_1$ and $A_2$ refer to the two scores for an
arbitrary subject in the experiment. Then we can refer to the parameter of
interest as $m = \mathrm{median}(A_2 - A_1)$.

With the bootstrap method, it is very easy to adapt our earlier code to work
for $m$. The statistic we want to use this time is the sample median of the
score differences.

***Exercises 10--14***

10. Calculate a point estimate of $m$.

11. Add extra code to your answer for Exercise 7 (and re-run it) so that you
    can use the same bootstrap samples to generate a CI for $m$. You simply
    need to calculate the new statistic and store it.

12. Calculate a 95% confidence interval for $m$ from your bootstrap samples,
    using the percentile bootstrap method.

13. Recreate the plot from Exercise 9, but now for inferring $m$. This time,
    you should only show the bootstrap-based CI.

14. Describe the approximate sampling distribution. What interesting features
    do you notice?

## Standard deviation of the differences

For our final step, we will infer the standard deviation of the pairwise
difference in scores, $\sigma = \mathrm{sd}(A_2 - A_1)$.

**Exercises 15--16**

15. What would be an appropriate statistic to use for this parameter?

16. Repeat Exercises 10--14 to perform inference for $\sigma$ using your chosen
    statistic.

# More information about functions

If you want to learn more about functions in R, the
[Functions](https://r4ds.hadley.nz/functions.html) chapter in the *R for Data
Science* book is a useful resource.
