---
title: "Week 6, Notes & Exercises"
subtitle: "Statistical Thinking (ETC2420 / ETC5242)"
date-format: "[Semester 2,] YYYY"
date: "2025"
toc: true
format:
  html:
    embed-resources: true
---

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(broom)
library(gridExtra)
library(MASS)
```

In this tutorial we will use maximum likelihood to estimate parameters for
various models, using simulated and real data. To quantify uncertainty, we
obtain both the CLT- and bootstrap-based confidence intervals.

# Normal distribution

## Generate the data

First, set the seed and then simulate a random sample of size $n = 36$ from
$\mathrm{N}(\mu = 10, \sigma^2 = 4)$.

```{r}
set.seed(24205242)
x_norm <- rnorm(n = 36, mean = 10, sd = 2)
```

Let's visualise the sample using the `dataplot()` function we wrote in the Week
5 tutorial. Suppose we saved the code for that function in a new file called
`dataplot.R`. We could then easily use it here as follows:

```{r}
source("dataplot.R")
dataplot(x_norm, bins = 30) +
  ggtitle("Histogram of simulated sample from N(10, 4)") +
  xlab("Observations (simulated)")
```

Notice the use of the `source()` function to load the code for the `dataplot()`
function. This is a quick and convenient way to re-use existing code without
needing to make multiple copies.

An alternative would be to copy and paste the code into the current document. A
downside of that approach is that if you notice an error in the code, then you
have to find all the places where you used it and fix it in each one. With the
`source()` approach, we only have a single copy of that code and so only need
to fix it once.

***Exercise 1***

1.  Create the file `dataplot.R` and put the definition of the `dataplot()`
    function inside it.

## Point estimates

Use the `fitdistr()` function (from the `MASS` package) to estimate the
parameters assuming a normal distribution.

```{r}
normal_fit <- fitdistr(x_norm, "normal")
normal_fit |> tidy()
```

Note that we are pretending we don't know the true parameter values, as if we
had collected these data and want to fit a normal distribution. This means
estimating the values of the parameters, $\mu$ and $\sigma$. Therefore there
are 2 parameters, so $$\theta =
\begin{pmatrix}      \mu  \\      \sigma  \end{pmatrix}, \quad
\hat{\theta}_\text{MLE} =
\begin{pmatrix} \hat{\mu} \\ \hat{\sigma} \end{pmatrix}.$$

The `fitdistr()` function can fit various probability distributions, estimating
their parameters using the maximum likelihood method using input data.

Have a look at the `fitdistr()` help page to find out more. Which models can it
fit? What does the output look like?

Let's explore the `normal_fit` object produced above. Look at what follows from
each of the print statements below. What does each represent? What is the
fitted population distribution?

```{r}
normal_fit$estimate
normal_fit$sd
normal_fit$vcov
normal_fit$loglik
normal_fit$n
```

We can see that that `normal_fit` is an object of class `fitdistr` which is a
list with 5 components:

-   `estimate` contains the MLEs of the parameters, ($\hat{\mu}, \hat{\sigma}$)
-   `sd` contains the standard errors (estimated standard deviations) of the
    parameter estimates
-   `vcov` contains the estimated variance-covariance matrix[^1] of the MLE
-   `loglik` contains the log-likelihood function evaluated at the MLE
    estimates
-   `n` contains the number of observations used in the fit.

[^1]: This is the equivalent quantity to a variance but for a multivariate
    probability distribution. The entries along the diagonal of this matrix are
    the estimated variances of the estimator for each parameter. The
    off-diagonal elements are estimated covariances. We won't work directly
    with these quantities in this unit, but it is helpful to know what they
    are.

When we estimate with `fitdistr()`, it returns point estimates as well as
standard errors. That allows us to get CLT-based confidence intervals for all
parameters, using the asymptotic distribution of MLEs.

Notice that this means we can now get a CLT-based CI for $\sigma$, whereas
previously we only knew how to do it for $\mu$.

## Confidence intervals

We now calculate 95% confidence intervals for the parameters using both the CLT
and the bootstrap.

***Exercises 2--4***

2.  Use the output from `fitdistr()`, along with appropriate quantiles from a
    normal distribution, to calculate approximate 95% confidence intervals for
    $\mu$ and $\sigma$. Use the fact that we know the MLE is asymptotically
    normally distributed.

3.  Use the bootstrap to calculate a 95% confidence interval for each
    parameter. Use $B = 5000$ bootstrap samples.

    Some code to get you started is below, but you'll need to change every
    instance of "`AAA`" with some suitable R code. Note that because we have
    two parameters to estimate, from each sample we need to store two values
    rather than just one. To do that, we need to use a matrix rather than a
    vector.

```{r}
#| eval: false
# Initialise values for doing the bootstrap.
B <- AAA
normal_boot <- matrix(nrow = B, ncol = AAA)

# Generate the bootstrap samples and calculate statistics.
for(b in 1:B) {
  AAA <- sample(AAA)
  normal_boot[b,] <- AAA
}

# Calculate CIs.
normal_boot_ci_mu    <- quantile(AAA, c(0.025, 0.975))
normal_boot_ci_sigma <- quantile(AAA, c(0.025, 0.975))

# Look at the values.
normal_boot_ci_mu
normal_boot_ci_sigma
```

For the CI calculation, notice that we calculate the lower and upper bounds at
the same time. However, we need to do it separately for each parameter, by
taking the appropriate column from the `normal_boot` matrix.

4.  Visualise the approximate sampling distributions, as given by the bootstrap
    samples, and include vertical lines to show the bounds of the 95%
    confidence intervals. (This is similar to what we did in Week 5.)

# Beta distribution

Repeat the work from the previous section, but now simulate a random sample of
size $n = 100$ from $\mathrm{Beta}(\alpha = 2, \beta = 4)$.

***Exercises 5--10***

5.  Simulate the data. Remember to set a seed to make it reproducible.

6.  Visualise the sample using an appropriate plot.

7.  Calculate MLEs and standard errors for the parameters, $\alpha$ and
    $\beta$.

    You might encounter an error when using `fitdistr()` if you merely copy
    what we did for the normal distribution above.

    The `fitdistr()` function requires the user to specify starting points for
    the optimisation algorithm (for maximising the log-likelihood) when working
    with the beta distribution. We suggest initialising the algorithm with
    $\alpha = 1$ and $\beta = 1$, although other positive values of $\alpha$
    and $\beta$ should work fine.

    Have a look at the help page for `fitdistr()` to find out how to specify
    these values.

8.  Calculate a CLT-based 95% confidence interval for each parameter.

9.  Use the bootstrap method to calculate 95% confidence intervals for the
    parameters.

10. Visualise the approximate sampling distributions, as given by the bootstrap
    samples, and include vertical lines to show the bounds of the 95%
    confidence intervals.

# Poisson distribution

Repeat the work from the previous section, but now simulate a random sample of
size $n = 75$ from $\mathrm{Poisson}(\lambda = 5)$.

***Exercises 11--16***

11. Simulate the data. Remember to set a seed to make it reproducible.

12. Visualise the sample using an appropriate plot.

13. Calculate the MLE and standard error of $\lambda$.

14. Calculate a CLT-based 95% confidence interval for $\lambda$.

15. Use the bootstrap method to calculate a 95% confidence interval for
    $\lambda$.

16. Visualise the approximate sampling distribution, as given by the bootstrap
    samples, and include vertical lines to show the bounds of the 95%
    confidence interval.

# CBT data

Load the CBT data and calculate the pairwise score differences.

```{r}
#| message: false
CBT <- read_csv("data/CBT.csv")
x_diff <- CBT$score2 - CBT$score1
```

## Estimating parameters

***Exercises 17--22***

17. Visualise the pairwise score differences with an appropriate plot.

18. What would be an appropriate distribution to fit to the score differences?

19. Using the distribution you chose above, fit it to the data using
    `fitdistr()`.

20. Calculate 95% confidence intervals for the parameters of the distribution
    you chose. Use both a CLT- and a bootstrap-based approach.

21. Visualise the approximate sampling distributions, as given by the bootstrap
    samples, and include vertical lines to show the bounds of the 95%
    confidence intervals.

22. We will now compare these estimates with those from the Week 5 tutorial.
    Are the inference results based on the MLE similar to the results from Week
    5? If any are different, can you explain why?

    (Before doing this exercise, we need to get MLEs using a normal
    distribution. If in the previous exercises you chose a distribution other
    than a normal distribution, then repeat them using a normal distribution.)

## Evaluating the model fit

We can use a QQ plot to assess whether a normal distribution is an adequate
model for the score differences.

We plot the data ("sample quantiles") against the theoretical quantiles from a
normal distribution. If most of the points are close to a straight line, then
we can be satisfied that a normal distribution is a satisfactory model.

```{r}
params_norm <- fitdistr(x_diff, "normal")$estimate
ggplot(tibble(x = x_diff)) +
  aes(sample = x) +
  stat_qq(distribution = qnorm, dparams = params_norm) +
  stat_qq_line(distribution = qnorm, dparams = params_norm) +
  theme_bw() +
  xlab("Theoretical quantiles (normal distribution)") +
  ylab("Sample (CBT score difference)")
```

***Exercises 23--25***

23. What can you conclude from the QQ plot?

24. Create another QQ plot, this time for assessing whether a t-distribution is
    an adequate model.

    Note: when you use `fitdistr()` for a t-distribution, it fits a 3-parameter
    version of this model. You only need to use the estimate of the `df`
    parameter when you create the QQ plot.

25. Do you think the t-distribution fits the data better?

## Confidence bands for QQ plots

The QQ plots above show a single set of points, based on the sample. It can
sometimes be hard to judge whether the points are 'close enough' to the
reference line to believe the model is adequate, or whether they 'deviate
enough' to cast doubt on the assumed model.

This problem is similar to only relying on a point estimate and not having any
sense of the underlying uncertainty in the estimate. We can improve our QQ
plots by adding a 'confidence band' to visualise the uncertainty. (A confidence
band is like an 'error bar', but for a whole line rather than a single point.)

We will use the [qqplotr](https://cran.r-project.org/package=qqplotr) package
to plot QQ plots with confidence bands. If you don't have this package
installed, you can install it as usual with the following command:

```{r}
#| eval: false
install.packages("qqplotr")
```

Then load it:

```{r}
library(qqplotr)
```

The package's [website](https://aloy.github.io/qqplotr/) and
[vignette](https://aloy.github.io/qqplotr/articles/introduction.html) show how
to use the package, but here's a brief example:

```{r}
ggplot(tibble(x = x_diff)) +
  aes(sample = x) +
  stat_qq_band(distribution = "norm", dparams = params_norm) +
  stat_qq_line(distribution = "norm", dparams = params_norm) +
  stat_qq_point(distribution = "norm", dparams = params_norm) +
  theme_bw() +
  xlab("Theoretical quantiles (normal distribution)") +
  ylab("Sample (CBT score difference)")
```

The package provides a new `stat_qq_band()` function (which draws the
confidence band), as well as its own `stat_qq_line()` and `stat_qq_point()`
functions (to draw the other elements of the QQ plot).

The package has several options for how to calculate the confidence band. The
details are somewhat advanced and we won't cover them in this unit. Also we
will just use the default settings.

The key aspects that you need to know are the following:

-   The confidence band shows the region where we expect most of the points to
    lie if the assumed model is indeed correct. It is centered on the reference
    line and 'flares out' away from the line in both directions. The region is
    wider on either end of the plot and narrower in the middle, which reflects
    the fact that the more extreme quantiles are 'noisier' and thus have
    greater uncertainty.

-   We don't necessarily expect **all** of the points to lie inside the band.
    Some might stray outside by chance, even if the assumed model is true. The
    more points there are, the more likely that a few of them will be outside
    by chance.

The confidence band gives us an easier way to judge whether the points are
close enough to the reference line. If the points lie completely within the
confidence band, then the assumed model is probably adequate for these data.
Otherwise, the more that the points deviate out of the band, the stronger the
evidence against the assumed model.

One difference compared to the standard `ggplot2` commands is how to specify
the assumed distribution using the `distribution` argument. When we use the
standard `ggplot2` commands, we pass a **function** that defines the inverse
cdf (`distribution = qnorm`). When using the `qqplotr` commands, we instead
pass the **name** of the set of functions without the usual d/p/q/r prefix
(`distribution = "norm"`). The reason for this different convention is that
`qqplotr` needs to access to the whole family of probability functions, not
just the inverse cdf.

***Exercises 26--27***

26. What can you conclude from the above plot?

27. Repeat this for the t-distribution.
