---
title: "Week 4, Notes & Exercises"
subtitle: "Statistical Thinking (ETC2420 / ETC5242)"
date-format: "[Semester 2,] YYYY"
date: "2025"
toc: true
format:
  html:
    embed-resources: true
---

```{r}
#| label: setup
#| include: false
#| warning: false
#| error: false
#| message: false
knitr::opts_chunk$set(message = FALSE,
                      echo    = TRUE,
                      eval    = TRUE,
                      fig.height = 3,
                      fig.width  = 7)
```

```{r}
#| eval: true
#| echo: false
#| warning: false
#| message: false
library(tidyverse)
library(kableExtra)
library(broom)
library(ggthemes)
theme_set(theme_minimal(base_size = 12))
```

# Gender discrimination study

Consider the case study example discussed in the lecture. The research question
we hope to answer is, "Are females discriminated against in promotion decisions
made by male managers?"

The participants in this study were 48 male bank supervisors attending a
management institute at the University of North Carolina in 1972. They were
asked to assume the role of the personnel director of a bank and were given a
job applicant resume to judge whether the applicant should be promoted to a
branch manager position. The files given to the participants were identical,
except that half of them indicated the job candidate was male and the other
half indicated the job candidate was female. These files were randomly assigned
to the supervisors to make their judgement.

The *gender* (male or female) associated with the assigned file and the
resulting promotion *decision* (to promote or not promote) was recorded. Using
the results of the study, we would like to evaluate if females are unfairly
discriminated against in promotion decisions.

We are interested in testing the upper one-sided hypothesis test:
$$H_0\colon p_{M} - p_{F} = 0 \quad \text{versus} \quad
  H_1\colon p_{M} - p_{F} > 0,$$ where ${p}_{M}$ and ${p}_{F}$ denote the
population proportions of male and female job applicants promoted,
respectively.

In this tutorial we will do two approximate versions of the test, using a
CLT-based test and a permutation test based on permutations of the original
data.

## Data wrangling

***Exercises 1--7***

1.  Create a tibble named `GDS` (abbreviation for "Gender discrimination
    study") that records the data from this study, with one row for each study
    participant. That means you should have 48 rows in total.

2.  Write code that will take `GDS` and count how many of each combination of
    *gender* and *decision* were observed in the study. The output should be a
    tibble that has one row for each such combination, and a column to show the
    count for each one. (Hint: use the functions `group_by()` and `tally()`.)

3.  Write code that will restructure this into a $2 \times 2$ table. (Hint: use
    the `pivot_wider()` function.)

4.  Write further code that will create output that is similar to the following
    table. Save your tibble into a variable called `GDSS4`, which we will use
    again later on. (Hint: use the `mutate()` and `select()` functions. When
    you do so, you can use backticks to quote the value `not promoted` to
    ensure it is treated as a single value when passed as an argument to those
    functions.)

```{r}
#| echo: false

# This code chunk is simply to show you an example of the desired output.
# You need to write code that will generate the same output, but by starting
# with the tibbles from  earlier exercises and manipulating them with
# appropriate R commands.

example_table <- tibble(gender         = c("male", "female"),
                        promoted       = c(21, 14),
                        "not promoted" = c(3, 10),
                        total          = c(24, 24),
                        prop           = c(0.875, 0.583))
example_table |>
  kable() |>
  kable_styling(latex_options = "hold_position")
```

5.  Using the output from Exercise 3 (see above), write code that will turn it
    back into "long" format, so it matches the output from Exercise 2. (Hint:
    use the `pivot_longer()` function.)

6.  What does the table from Exercise 4 (see above) tell us? Compare the
    proportions of men and women who were promoted, and calculate the
    difference. Does it match our expectations?

7.  Is this evidence of discrimination?

## Visualise the data

Another way to present the results would be using a plot. Since the data are
categorical, we can use a bar plot. The code chunk below produces a single plot
comprised of two bar plots, each showing the counts of the `decision` variable
for each gender, positioned side-by-side. (Before you can run it, you need to
complete Exercise 1, in which you create the tibble `GDS`.)

```{r}
#| eval: false
ggplot(GDS, aes(x = decision, fill = decision)) +
  geom_bar() +
  facet_wrap(~ factor(gender), nrow = 1)
```

Notice the following:

-   The `geom_barplot()` function provides a summary of the tidy `GDS` tibble
    (i.e., it used the individual-level data, rather than a summarised version
    where the counts were already provided).

-   By using `facet_wrap(~ factor(gender))`, the two bar plots share the same
    vertical (y-)axis, and we automatically get titles for each panel, which
    correspond to the different values of the `gender` variable.

-   See also `facet_grid()`, as an alternative to `facet_wrap()`.

***Exercise 8***

8.  What happens if the `facet_wrap()` function is removed from the code chunk
    shown above? Which plot do you think highlights the different proportions
    best? Is the plot now evidence for discrimination?

## Comparison of proportions

Let the difference between the proportions of interest be $d = p_{M} - p_{F}$.

Let the estimators of these parameters be the natural sample equivalents. That
is, let $\hat{p}_{M}$ and $\hat{p}_{F}$ be the sample proportions of male and
female job applicants promoted, respectively, and let $\hat{d} = \hat{p}_{M} -
\hat{p}_{F}$.

***Exercise 9***

9.  Let `GDSS4` be the tibble that formed the basis for the output of Exercise
    4 (see above). The code chunk below provides three different ways of
    computing $\hat{d}$ from this tibble. While each one provides the same
    numerical value, which expression do you think should be preferred, and
    why?

```{r}
#| eval: false
d.hat <- GDSS4$prop[1] - GDSS4$prop[2]
d.hat <- GDSS4$prop[GDSS4$gender == "male"] - GDSS4$prop[GDSS4$gender == "female"]
d.hat <- (GDSS4 |> pull(prop))[1] - (GDSS4 |> pull(prop))[2]
```

# How do we test for discrimination?

We wish to test if the sample supports the hypothesis that the difference in
proportions is 0 versus it being greater than 0. We will use the observed
difference in proportions ($\hat{d}$) as the test statistic. To do so, we need
a way to calculate and work with the sampling distribution of $\hat{d}$. We
explore two choices: a CLT-based approach and a permutation approach.

## The CLT-based test

Here we use the function `prop.test()` to do the relevant traditional CLT-based
test.

This function is not part of the tidyverse and needs a simpler format for the
input (it takes vectors or matrices, rather than tibbles). See the code below.

Note that the default for `prop.test()` is to do a two-sided test. Since we
want a one-sided test, we need to specify the `alternative` argument to tell it
which one-sided alternative hypothesis we want.

```{r}
#| eval: false
GDSS4 |>
  select(promoted, `not promoted`) |>
  as.matrix() |>
  prop.test(alternative = "greater")
```

***Exercise 10***

10. Report the conclusion of the hypothesis test on the basis of the above
    output, and comment on the strength of evidence against $H_0$.

## The permutation test

The calculations to implement the permutation test in this setting are shown in
the code chunk below (with some missing parts that you need to fill in, see the
exercises).

We want to "break" the association between `promoted` and `gender` by permuting
(shuffling) one of the columns. Read through the code below carefully.

The `sample()` function does the shuffling, and the loop does this many ($B$)
times. Each of these $B$ shuffles is a hypothetical version of the data
assuming that $H_0$ were true. Therefore, we get $B$ simulated values from the
sampling distribution of $\hat{d}$ if $H_0$ were true (we often refer to this
simply as the *null distribution* of $\hat{d}$).

To visualise the approximate null distribution of $\hat{d}$, we can plot a
histogram of the B simulated values.

We can also use the simulated values to calculate approximate p-values, by
counting what proportion of the simulated values are as or more extreme than a
given observed value of $\hat{d}$.

```{r}
#| eval: false

# Set the number of replications for test.
B <- 1000

# Set up an empty vector of length B, to be used to store the
# replicated "d.hat" values after shuffling.
Bdhat <- numeric(B)

# Initialise permutation sample with the original data.
# We will update this inside the loop with each permutation.
RGDS <- GDS

# ...comment left blank, see the exercise below...
set.seed(2025)

# Use for-loop to generate B replications of the test statistic.
# All commands inside the curly braces are repeated B times,
# with the variable r taking values in the sequence 1:B in turn.
for (r in 1:B) {

  # For iteration r, shuffle the gender variable in the data file.
  RGDS <- RGDS |> mutate(gender = sample(RGDS$gender))
  
  # Calculate the summary table for the shuffled tibble, in the
  # same way as for Exercises 2 to 4.
  RGDSS3 <- RGDS |>
    ((...add your code here...))

  # Calculate the "d.hat" for the shuffled data, and
  # save it as the r-th element of the Bdhat vector.
  Bdhat[r] <- ((...add your code here...))

  # Close the for-loop with a curly brace.
}

# Proportion of Bdhat at least as big as d.hat.
# This is the approximate p-value.
pval <- sum(Bdhat >= d.hat) / B
```

***Exercises 11--13***

11. Write the missing R code.

12. What does the `set.seed(2025)` command do? Why is it included in the code
    for the permutation test?

13. Set $B = 5000$ and re-run the code (it will take longer to run). Does the
    answer change?

## Visualise the sampling distribution and the p-value

Since we simulated values from the sampling distribution, we can plot them.
Consider the following code chunk and resulting plot.

```{r}
#| eval: false
Bdhat |>
  as_tibble() |>
  mutate(r = 1:B) |>
  ggplot(aes(value)) +
  geom_histogram(fill = "blue", alpha = 0.7) +
  xlab(expression(hat(d)^"[r]")) + 
  ggtitle(expression(paste("Approximate null distribution of ",
                           hat(d) == hat(p)[M] - hat(p)[F])))
```

***Exercises 14--17***

14. Modify the above plot to add a vertical line at the observed value of
    $\hat{d}$.

15. What does this plot represent?

16. Report the p-value.

17. Would you say that we have proved that the managers discriminate against
    women?
