---
title: "Week 11, Notes & Exercises"
subtitle: "Statistical Thinking (ETC2420 / ETC5242)"
date-format: "[Semester 2,] YYYY"
date: "2025"
toc: true
format:
  html:
    embed-resources: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE)
```

This week we continue our exploration of the Lego dataset (see Week 10), now
focusing on prediction.

First we load some R packages:

```{r}
#| message: false
library(tidyverse)
library(openintro)
library(broom)
```

# Revision

We start with some revision from last week...

***Exercises 1--2***

1.  Load the Lego data and prepare it the same way as last week (add the
    `is_city` variable, etc.).

2.  Which model was selected as the best model for prediction in last week's
    exercises?

# Estimating prediction performance

We now explore some measures of prediction performance, based on using only the
original data (the same as we used for fitting the models, i.e. the 'training
data').

In the following exercises, you need to use the following three models:

a)  The 'best' model (see Exercise 2)
b)  The full model (which includes all predictors)
c)  The model with only the predictors `pieces` and `is_city`.

***Exercises 3--5***

3.  Calculate the AIC for each of the three models.

4.  Calculate the LOOCV estimate of the prediction error for the three models.

    Hints:

    -   You don't need to implement cross-validation directly. You can use the
        formula presented in the lecture.

    -   If you write a function that calculates LOOCV for any given model, you
        can simply apply it to all three models.

5.  Compare the AIC and LOOCV values for the above models and describe what you
    see.

# Prediction in a new dataset

We now wish to use the best model to predict the sale price for some Lego sets
that were not in our original sample.

Run the following code to get a new dataset to work with:

```{r}
#| eval: false
desired_rows <- c(82, 118, 159, 302, 420, 624, 802, 972, 1124, 1258)
lego_new <- lego_population[desired_rows, ]
```

***Exercises 6--7***

6.  Repeat the process you used in Exercise 1 to replace missing values with
    suitable values.

7.  Using the best model, predict the sale price for these new Lego sets.
    Include a 95% prediction interval.

Do you notice any problems with these predictions?

The prediction intervals often give **negative** values, which does not make
sense for a sale price. This is a deficiency of the model, and is due to the
fact that the model imposes no constraints on the response variable (we assume
it is normally distributed around the conditional mean given by the model
equation).

The fact that sale prices should be positive numbers also leads to another
visible defect in the model: the residuals are heteroscedastic (as we observed
last week; see Exercise 9 in Week 10).

One way to try to fix both problems is to log-transform the sale prices before
any modelling. That would ensure that only positive prices are possible. It
will also use a scale that makes larger prices closer together, while spreading
out the smaller prices, hopefully making them have similar variance across the
whole set of prices.

Often this works well and gives better models. However, it can also sometimes
be too strong a transformation, and using something like a square root
transformation can be a good compromise.

When we apply any transformations, the data usually look quite different to
before, so we should repeat all of the modelling steps and diagnostic checks.
You can try doing a complete version of this on your own; for today, we will
just do a more limited set of steps, as detailed in the following exercises.

***Exercises 8--9***

8.  Apply each of the following transformations to the sale price, re-fit the
    model and re-do the predictions:

    a)  Log
    b)  Square root

    What can you say about the predictions?\
    Also, compare them to the reported sale prices that are in `lego_new` and
    comment.

9.  Plot the residuals against fitted values for each version of the model.
    What can you conclude?
