---
title: "Week 7, Notes & Exercises"
subtitle: "Statistical Thinking (ETC2420 / ETC5242)"
date-format: "[Semester 2,] YYYY"
date: "2025"
toc: true
format:
  html:
    embed-resources: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

For this week, we take our first steps in Bayesian inference. In the first
part, we revise some probability and practise using Bayes' Theorem. In the
second part, we learn more about the application of Bayes' Theorem and practise
using it in a context where the unknown variable takes on only a finite number
of values. In the third part, we consider a scenario where the unknown variable
can take an infinite number of possible values. For simplicity, we focus only
on conjugate situations.

This week is a good opportunity to practise including mathematical symbols in
your Quarto files, and combine them with doing the calculations using R. A
useful resource for LaTeX math symbols can be found at [this
page](https://www.caam.rice.edu/~heinken/latex/symbols.pdf).

# Food hygiene

A food hygiene inspection of Melbourne restaurants is carried out routinely.
The outcome of each inspection is either a pass or a fail. The inspectors are
not always reliable at picking up all food hygiene issues. From past
experience, it is known that the probability of a pass if food hygiene is
satisfactory is 0.9, and if it is unsatisfactory is 0.3.

***Exercises 1--3***

1.  Fast food restaurants are known to have unsatisfactory standards about 20%
    of the time. If an inspection of such a restaurant results in a fail, what
    is the probability it has unsatisfactory food hygiene standards?

2.  Fine dining restaurants are known to have unsatisfactory standards only
    about 2% of the time. If an inspection of such a restaurant results in a
    fail, what is the probability it has unsatisfactory food hygiene standards?

3.  Draw tree diagrams to represent this problem.

# Is that a typo? Spell checking using Bayes' Theorem

Have you ever wondered how an internet search engine decides when to correct
the spelling of a word?[^1] The answer is...by using Bayes' Theorem![^2] Let's
see how it works.

[^1]: Much of this question has been taken from *Bayesian Data Analysis*,
    Gelman A, Carlin JB, Stern HS, Dunson DB, Vehtari A and Rubin DB. 3rd
    edition, Chapman & Hall/CRC texts in statistical science, 2014.

[^2]: The actual spell checking algorithms are much more complex than described
    here. But the basic principle is right. For more information, see the
    following blog post by Peter Norvig:
    <http://www.norvig.com/spell-correct.html>.

Do a search on the word 'radom' using Google. What information do you find?

## Now imagine that you are the computer search engine...

When you type a word like 'radom' into a Google search, how does Google know
whether you really intended to type 'radom' or whether instead you made a
spelling error? It seems to consider that you may have intended to type a
different word, such as 'random' or 'radon'.

In statistics and machine learning, a *language model* is a probability
distribution defined over sequences of words. Simple language models are
defined for individual words found in a nominated 'universe of words' -- a
collection usually obtained from scans of historical texts encompassing at
least a million words. The model then defines the probability[^3] of each word
(meaning the chance it will occur again in a new sentence) as a proportion of
times that word appears in the nominated 'universe of words'. Some language
models cover a wide variety of topics, while others may be more context
specific. For example, the 'universe of words' file may relate specifically to
science, so words that occur more often in scientific discussions would be
associated with a higher probability than they might be in a more general
language model. More complex language models construct probabilities for
sequences of words, and again can be context specific.

[^3]: Some would argue this is really an *estimate* of the relevant
    probability. Alternatively, it may be taken is as an assessment of *prior
    belief* about the next likely intended word to appear in the search engine.

We will work with a simple language model, and focus on only three words. While
a simplification, it provides a nice introduction to the use of Bayes' Theorem
in a machine learning application.

## The language model

Here we construct the language model probabilities from (past) relative
frequencies of the three words available from an available universe of words.

Let the variable $W$ represent the *word that was intended to be typed* on a
given occasion. The three possible values of $W$ in this setting are given by:

-   $W_1 =$ 'random'
-   $W_2 =$ 'radon'
-   $W_3 =$ 'radom'

***Exercise 4***

4.  How common do you think each of these three words in the English language?
    (If you are not familiar with the word 'radon', you can ask Google for a
    definition.)

Now consider the available relative frequencies, denoted by $RF_i$, for each
$W_i$, as shown in @tbl-relfreq. The word 'random' occurs much more frequently
than does the word 'radon', which in turn occurs much more frequently than the
word 'radom'. (Does this seem sensible to you?)

| $W_i$  | $RF_i$                |
|:-------|:----------------------|
| random | $7.6 \times 10^{-5}$  |
| radon  | $6.05 \times 10^{-6}$ |
| radom  | $3.12 \times 10^{-8}$ |

: Relative frequency ($RF_i$) for each intended word ($W_i$). {#tbl-relfreq}

***Exercise 5***

5.  With the words 'random', 'radon' and 'radom' being the only words under
    consideration, construct marginal probabilities for each word as implied by
    the relative frequencies shown in @tbl-relfreq.

## An observation

Now suppose Google's search engine has been given the search request, 'radom',
which is a word it recognizes. This is an *observation*, which we will denote
as $y$, i.e. $y = \text{`radom'}$.

Google's search engine is programmed to consider whether the typed word is the
word that was intended or if a different word was intended and a typographical
error occurred. Bayes' Theorem is used to update the marginal probabilities
implied by the information in @tbl-relfreq, to incorporate the observation of
$y =
\text{`radom'}$.

## An error model for $y = \text{`radom'}$

Now to actually use Bayes' Theorem, we need to be able to combine the new
information about the observation $y$ with the marginal probabilities we have
for each $W_i$. In this language context, an *error model* is used as it
identifies the chance that $y$ will occur (as a *mis-typed* word), when in fact
the true intended word is $W_i$.

Google has a model of spelling and typing errors[^4], which provides the
conditional probabilities shown in @tbl-errormodel. (Do you think these error
model probabilities seem reasonable?)

[^4]: Such a model may have been developed from a study where people were
    followed up after writing emails, so that the intended meaning of words
    that might have seemed incorrect in those emails could be identified.

| $W_i$  | $\Pr(y = \text{`radom'} \mid W_i)$ |
|:-------|:-----------------------------------|
| random | $0.00193$                          |
| radon  | $0.000143$                         |
| radom  | $0.975$                            |

: From Google's model of spelling and typing errors, in the case when
$y = \text{`radom'}$. {#tbl-errormodel}

As the observation $y = \text{`radom'}$ is fixed, we are viewing $\Pr(y =
\text{`radom'} \mid W_i)$ as a function of $W_i$, with $i \in \{1,2,3\}$ and
the data fixed at the observed value $y = \text{`radom'}$. Recall that this
function is called the *likelihood* function.

***Exercise 6***

6.  Should the probabilities in @tbl-errormodel sum to one? Why or why not?

## Updated probabilities

| $W_i$  | $\Pr(W_i \mid y = \text{`radom'})$ |
|:-------|:-----------------------------------|
| random |                                    |
| radon  |                                    |
| radom  |                                    |

: Conditional distribution of $W_i$ given $y$. {#tbl-posteriors}

***Exercises 7--9***

7.  Use Bayes' Theorem to complete the values for @tbl-posteriors. State your
    answers to at least four decimal places.

    Note that your updated probabilities should combine *both* of the following
    facts:

    -   That 'radom' occurs relatively infrequently, according to the given
        language model, compared to the word 'random'.

    -   Depending on the intended word, the chance of actually typing 'radom'
        can happen with different probabilities.

8.  Should the probabilities in @tbl-posteriors sum to one? Why or why not?

9.  If you had used the relative frequencies $RF_i$ (from @tbl-relfreq) in
    place of the normalised probabilities $\Pr(W_i)$ in Bayes' Theorem, would
    your resulting conditional probabilities have the same values? Why or why
    not?

# Working with Bernoulli trials

## The data model

Consider an experiment where a random sample of $n$ observations, each taking
one of two possible (binary) outcomes, with $$u_{i} =
\begin{cases}
1, & \text{if success,} \\
0, & \text{if failure.}
\end{cases}$$

These are *Bernoulli trials*: each of these random outcomes are independent and
have the same probability of a success, which we will denote here as $\theta$.
The likelihood function corresponding to $U_1, U_2, \dots, U_n \mid \theta
\overset{\mathrm{iid}}{\sim} \mathrm{Bernoulli}(\theta)$ trials is $$L(\theta)
= \prod_{i=1}^{n}\theta^{u_{i}}(1-\theta)^{1-u_{i}}
= \theta^{\sum_{i=1}^{n}u_{i}}(1-\theta)^{n-\sum_{i=1}^{n}u_{i}}.$$ Rather than
thinking of modelling each of the individual $U_i=u_i$ outcomes as a single
$\mathrm{Bernoulli}(\theta)$ random variable, we can consider the sum of $n$
such $\mathrm{Bernoulli}(\theta)$ outcomes, $X=\sum_{i=1}^{n}U_{i}$. In this
case, $X \mid \theta \sim \mathrm{Binomial}(n,\theta)$ distribution, taking on
a value $x\in \{0, 1, \dots, n\}$. Once $X=x$ is observed, the likelihood
function associated with $X \mid \theta \sim
\mathrm{Bernoulli}(\theta)$ is given by
$$L(\theta) = \binom{n}{x}\theta^{x}(1-\theta)^{n-x}.$$

Note that the combinatorial coefficient $\binom{n}{x} = \frac{n!}{x! (n-x)!}$
does not depend on $\theta$.

## A beta prior for the binomial (or Bernoulli) likelihood

The $\mathrm{Beta}(\alpha, \beta)$ distribution is the prior conjugate to the
binomial likelihood function. Remember the probability density function (pdf)
for $\theta \sim \mathrm{Beta}(\alpha, \beta)$ is
$$f(\theta \mid \alpha, \beta)
= \frac{\Gamma(\alpha + \beta)}
       {\Gamma(\alpha) \Gamma(\beta)}
  \theta^{\alpha-1}(1-\theta)^{\beta - 1},
\quad \text{for } \theta \in (0,1), \; \alpha>0, \; \beta>0.$$ Also we know the
values of the prior mean and variance:

-   $\mathbb{E}(\theta)   = \frac{\alpha}{\alpha + \beta}$
-   $\mathrm{var}(\theta) = \frac{\alpha \beta}{(\alpha+\beta)^2(\alpha+\beta+1)}$

Probabilities for $\theta$ (e.g., $\Pr(\theta < 0.2)$) and quantiles associated
with the $\mathrm{Beta}(\alpha, \beta)$ distribution may be obtained using the
R functions `pbeta()` and `qbeta()`.

## Bayes' Theorem

Bayes' Theorem, stated for a continuous parameter and a model for continuous
data, says that if $X \mid \theta$ has pdf given by $f(x \mid \theta)$, and
$\theta$ has pdf $f(\theta)$, then the posterior pdf for $\theta \mid X = x$
satisfies $$f(\theta \mid x) =
\frac{f(x \mid \theta)~f(\theta)}
     {\int_{\Theta} ~f(x \mid \theta)~f(\theta)~d\theta}.$$

## Beta-binomial conjugate pair

If $\theta \sim \mathrm{Beta}(\alpha, \beta)$, for $0 < \theta < 1$, denotes
the prior distribution, and if $X$ given $n$ and $\theta$ has a
$\mathrm{Binomial}(n,\theta)$ distribution, then the posterior distribution for
$\theta$ when $X = x$ will be a $\mathrm{Beta}(\tilde{\alpha}, \tilde{\beta})$,
where $\tilde{\alpha} = \alpha + x$ and $\tilde{\beta} = \beta + (n - x)$.

The parameter we wish to infer is $\theta$. We also have parameters $\alpha$
and $\beta$, which are used to define the prior distribution. To distinguish
between these two different types of parameters, we usually refer to parameters
that define prior and posterior distributions (e.g., $\alpha$ and $\beta$) as
*hyper-parameters*.

***Exercises 10--17***

10. Create a function, named `beta_binomial()`, that takes as input the
    following values:

    -   $n =$ the sample size (the number of iid observations from
        $\mathrm{Bernoulli}(\theta)$)
    -   $x =$ the number of successes in the $n$ observations
    -   $\alpha =$ the first hyper-parameter of the $\mathrm{Beta}(\alpha,
        \beta)$ prior distribution (with default value = 1)
    -   $\beta =$ the second hyper-parameter of the $\mathrm{Beta}(\alpha,
        \beta)$ prior distribution (with default value = 1)

    and returns as output a list containing the elements:

    -   $\tilde{\alpha} =$ the first hyper-parameter of the
        $\mathrm{Beta}(\tilde{\alpha}, \tilde{\beta})$ posterior distribution
    -   $\tilde{\beta} =$ the second hyper-parameter of the
        $\mathrm{Beta}(\tilde{\alpha}, \tilde{\beta})$ posterior distribution

11. Now, suppose $\theta \sim \mathrm{Beta}(\alpha = 4, \beta = 4)$ when $X$ is
    modelled with a $\mathrm{Binomial}(n = 40, \theta)$ distribution, and $x =
    12$ is observed. Use your function to determine the posterior distribution.
    (Also, verify that you can get this same answer by manual calculation.)
    Explain what this distribution represents.

12. Visualise the prior and posterior pdfs, and the likelihood function.
    Superimpose these three curves together on the same plot.

    Hint: some R code to help you with the visualisation is below. You will
    need to provide the values of some of the variables, such as `alpha` and
    `Q1out`.

    ```{r}
    #| label: firstplot
    #| echo: true
    #| eval: false
    thetaxx <- seq(0.001, 0.999, length.out = 100)
    priorxx <- dbeta(thetaxx, alpha, beta)
    postxx  <- dbeta(thetaxx, Q1out$alpha_tilde, Q1out$beta_tilde)
    likexx  <- dbinom(xobs, size = n, prob = thetaxx)
    nlikexx <- 100 * likexx / sum(likexx)

    df <- tibble(theta                   = thetaxx,
                 `prior pdf`             = priorxx,
                 `normalised likelihood` = nlikexx,
                 `posterior pdf`         = postxx)

    df_longer <- df |>
      pivot_longer(-theta, names_to = "distribution", values_to = "density")

    p1 <- df_longer |>
      ggplot(aes(x = theta, y = density, colour = distribution)) +
      geom_line() +
      theme_bw()
    p1
    ```

13. Using the prior and posterior distributions from above, calculate both the
    prior probability $\Pr(0.2 < \theta < 0.4)$, and the posterior probability
    $\Pr(0.2 < \theta < 0.4 \mid x = 12)$.

14. Write a function, named `beta_meanvar()`, that takes as input the parameter
    values `alpha` and `beta`, corresponding to $\alpha$ and $\beta$,
    respectively, and returns a list containing the mean (named `mean`) and
    variance (named `var`) of the $\mathrm{Beta}(\alpha, \beta)$ distributions.

15. Use your function to calculate the mean and variance for both the $\theta
    \sim \mathrm{Beta}(\alpha = 4, \beta = 4)$ prior distribution, and for the
    corresponding posterior distribution for $\theta$ when $x = 12$.

16. Explain why the denominator in Bayes' Theorem above, once evaluated, is
    **not** a function of $\theta$.

17. Using the data $x = 12$ from $n$ Bernoulli trials, and the
    $\mathrm{Beta}(\alpha = 4, \beta = 4)$ prior, calculate the value of this
    denominator as follows:
    $$f(x) = \int_{\Theta} ~f(x \mid \theta)~f(\theta)~d\theta
           = \frac{f(x \mid \theta^{\ast})~f(\theta^{\ast})}
                  {f(\theta^{\ast} \mid x)},$$ for any
    $\theta = \theta^{\ast} \in (0,1)$.

    \[Hint: remember you can evaluate the likelihood using the `dbinom()`
    function, and you can evaluate both the prior and the posterior pdfs using
    the `dbeta()` function. As for $\theta^{\ast}$, any value in $(0,1)$ should
    give you the same answer, so just pick a few to check that you always get
    the same answer!\]
