---
title: "Week 8, Notes & Exercises"
subtitle: "Statistical Thinking (ETC2420 / ETC5242)"
date-format: "[Semester 2,] YYYY"
date: "2025"
toc: true
format:
  html:
    embed-resources: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

This week we continue learning about Bayesian inference. In the first part, we
build upon the Bernoulli trials scenario tackled at the end of last week. In
the second part, we look a simple 'A/B testing' example that mimics a common
scenario used in marketing analytics.

# Credibility factors for Bernoulli trials

The exercises in this section use the same model as in the **Working with
Bernoulli trials** section from last week's notes (Week 7).

We have $n$ Bernoulli trials, each taking one of two possible outcomes:
'success' or 'failure'. The data can be summarised by a single number $X$ which
is the number of successes. Therefore, $X \mid \theta
\sim \textrm{Binomial}(n, \theta)$, where $\theta$ is the probability of a
success on any given trial. For inferring $\theta$, we use a
$\mathrm{Beta}(\alpha, \beta)$ prior.

Let $z$ be the *credibility factor* for estimating $\theta$ by the posterior
mean, and $\hat\theta$ be the MLE for $\theta$. We have the following relation:
$$\mathbb{E}(\theta \mid x) =    z  \times \hat\theta +
                              (1-z) \times \mathbb{E}(\theta).$$

***Exercises 1--3***

1.  For each term in the above relation, express it only in terms of statistics
    and hyper-parameters.

2.  Take the `beta_binomial()` function you wrote last week and extend it to
    also output the credibility factor $z$.

3.  Verify the above relation for the case where we observe 12 successes out of
    40 trials and use a uniform prior. Use the `beta_binomial()` function you
    have just written above, together with the `beta_meanvar()` function you
    wrote last week.

# A/B testing

What is A/B testing?

Here's a brief description from the [Wikipedia
article](https://en.wikipedia.org/w/index.php?title=A/B_testing&oldid=1308290929):

> **A/B testing** (also known as **bucket testing**, **split-run testing** or
> **split testing**) is a user-experience research method. A/B tests consist of
> a randomized experiment that usually involves two variants (A and B),
> although the concept can be also extended to multiple variants of the same
> variable. It includes application of statistical hypothesis testing or
> "two-sample hypothesis testing" as used in the field of statistics. A/B
> testing is employed to compare multiple versions of a single variable, for
> example by testing a subject's response to variant A against variant B, and
> to determine which of the variants is more effective.

## A promotional email campaign

A/B testing is often used as part of email marketing. Consider an experiment
involving emails sent to 2000 different people independently sampled from a
very large customer database. All email messages contain a promotional offer to
buy a certain product with a discount, and are **identical** except for one
sentence:

-   Message A: "Offer ends this Saturday!"
-   Message B: "Offer ends soon!"

Half of the emails sent ($n_A = 1000$) include message A and the other half of
the emails ($n_B = 1000$) include message B. Those sent message A message
belong to **group A** and the remaining who received message B belong to
**group B**. There is no overlap between the two groups.

We define the *conversion rate* as the proportion of people (out of a given
group) who purchase a product. Let conversion rates for groups A and B be
$\theta_A$ and $\theta_B$, respectively.

We want to answer questions such as:

-   Which of the two messages, A or B, is more effective?\
    (In other words, which one results in a higher conversion rate?)

-   How much more effective is one message compared to the other?\
    (In other words, what is the distribution of the difference or ratio
    between $\theta_A$ and $\theta_B$?)

For some of the exercises given below, you will need to use the following
fact:[^1]

[^1]: This is a consequence of a more general result that is sometimes referred
    to as the 'double expectation' formula or the 'law of total expectation'.
    See the corresponding [Wikipedia
    article](https://en.wikipedia.org/wiki/Law_of_total_expectation) if you
    wish to find out more.

> If $X \mid \theta \sim \textrm{Binomial}(n,\theta)$, where $n$ is fixed and
> $\theta$ is random, then we can calculate the expected value of $X$ using
> $\mathbb{E}[X] = \mathbb{E}_{(X, \theta)}[X] = \mathbb{E}_{\theta} \left[n
> \theta\right] = n \mathbb{E}_{\theta} \left[\theta\right] = n
> \mathbb{E}[\theta]$.

The subscripts on the expectation operator, $\mathbb{E}$, refer to the random
variables over which the expectation (of the expression inside the square
brackets) is to be calculated. Note the similarity between the above result and
the corresponding formula if $\theta$ is considered fixed:
$\mathbb{E}[X] = n \theta$.

## Setting the prior

For analysing the data from this experiment, the company running the A/B test
has a lot of experience using these email marketing strategies and is happy to
use independent $\mathrm{Beta}(\alpha = 2,\beta = 26)$ prior distributions for
the two conversions rates.

***Exercises 4--6***

4.  According to the prior distribution, what is the expected number of people
    in group A who will take up their promotional offer?

5.  Prior to seeing any data, calculate a 95% probability interval $(l, u)$ for
    $n_A \theta_A$. In other words, we require some values $l$ and $u$ that
    satisfy $\Pr(l \leqslant n_A \theta_A \leqslant u) = 0.95$. (This is
    similar to a credible interval, but we are using the prior rather than the
    posterior.)

6.  According to the prior distributions for $\theta_A$ and $\theta_B$, and
    given $n_A = n_B = 1000$, what is the difference between the expected
    number of people from group A who will take up their promotional offer and
    the expected number of people from group B who will take up their
    promotional offer? Explain why.

## The experimental data

After running the A/B test, we observed that $x_A = 63$ customers purchased the
product after viewing message A and $x_B = 45$ customers purchased the product
after viewing message B.

***Exercise 7***

7.  Given the results of the experiment, what are the posterior distributions
    for the conversion rates for groups A and B, respectively?

## Comparing the conversion rates

Let $\Delta = \theta_B - \theta_A$ denote the difference between the two
conversion rates.

Although we can work out the posterior mean and variance of $\Delta$ fairly
easily, we do not know its full distribution. However, given what we know about
the posterior distributions of $\theta_A$ and $\theta_B$, it is easy to obtain
*draws* (realisations) from this distribution via simulation.

To use simulation, we can draw a large number, $R$, of values of
$\theta_A^{(r)} \overset{\mathrm{iid}}{\sim} f(\theta_A \mid x_A = 63)$ and
another $R$ values of
$\theta_B^{(r)} \overset{\mathrm{iid}}{\sim} f(\theta_B \mid x_B = 45)$. Then
we use these to compute $R$ realisations of
$\Delta^{(r)} = \theta^{(r)}_B - \theta^{(r)}_A$. We can approximate the
posterior distribution, $f(\Delta \mid x_A = 63, x_B = 45)$, using these
simulated $\Delta^{(r)}$ draws. For example, by visualising them or calculating
statistics of interest.

***Exercises 8--10***

8.  Given the results of the experiment, what is the (posterior) mean and
    variance of $\Delta$?\
    (Hint: Are the two posterior distributions independent? Why or why not?)

9.  Simulate a set of $R = 10,000$ draws of $\Delta$ from
    $f(\Delta \mid x_A = 63, x_B = 45)$. Use these to approximate each of the
    following quantities:

    a)  the posterior mean of $\Delta$,\
    b)  the standard deviation of $\Delta$, and\
    c)  the probability that the two conversion rates will differ by more than
        0.01, i.e. $$\Pr(\lvert \theta_B - \theta_A \lvert > 0.01 \mid x_A
        = 63, x_B = 45).$$

    Here is some R code to help you visualise the pdf of the posterior using
    your simulated values:

    ```{r}
    #| eval: false
    R <- 10000
    set.seed(24205242)

    # ...add your code here...

    ds <- tibble(delta = sampleD)
    ds |> ggplot(aes(x = delta, y = after_stat(density))) +
      geom_histogram(colour = "blue", fill = "blue", alpha = 0.4, bins = 100) +
      geom_density(  colour = "blue", fill = "blue", alpha = 0.4) +
      geom_vline(xintercept = c(0.01, -0.01), colour = "red") +
      ggtitle(expression(paste("Simulated posterior pdf of ", Delta))) +
      xlab(expression(Delta)) +
      theme_bw()
    ```

10. Of the non-Bayesian techniques we have learnt about thus far in this unit,
    which do you think would be appropriate to apply in this setting? How would
    you implement them and what type of inference/conclusions would they
    provide?
