---
title: "Week 2, Notes & Exercises"
subtitle: "Statistical Thinking (ETC2420 / ETC5242)"
date-format: "[Semester 2,] YYYY"
date: "2025"
toc: true
format:
  html:
    embed-resources: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo    = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      echo    = TRUE,
                      eval    = TRUE)
set.seed(24205242)

library(tidyverse)
```

# Introduction

In this tutorial we will practise doing probability calculations in R, using
simulation as a tool to answer tricky questions, modelling using probability
distributions, and using randomisation as part of designing a study.

# Working with probability distributions

R has functions for calculating with many different distributions, to get a
full list of those available in a standard R installation, run the following
command:

```{r}
#| eval: false
help(Distributions)
```

The functions for each distribution use a similar interface and conventions,
which we illustrate here first by looking at the normal distribution.

## Normal distribution

For most distributions, R provides four functions. The convention is that each
function has the same name with but a prefix to denote what the function does.
Here are the four functions for the normal distribution:

-   `dnorm()` -- calculates the probability density function (pdf)
-   `pnorm()` -- calculates the cumulative distribution function (cdf)
-   `qnorm()` -- calculates quantiles (inverse cdf)
-   `rnorm()` -- generates random numbers

Given a set of values, `dnorm()` returns the value of the pdf at each point.
The default is to work with a standard normal distribution ($\mu = 0$ and
$\sigma = 1$), but you can change this by specifying different function
arguments (see the help page for more info). Try the following and think about
what each one is calculating (discuss and compare your ideas with your
classmates):

```{r}
#| eval: false
dnorm(0)
dnorm(0) * sqrt(2 * pi)
dnorm(0, mean = 4, sd = 2)
dnorm(c(-1, 0, 1))
x <- seq(-20, 20, by = 0.1)
y <- dnorm(x)
ggplot() + aes(x = x, y = y) + geom_point()
y <- dnorm(x, mean = 2.5, sd = 0.5)
ggplot() + aes(x = x, y = y) + geom_point()
```

The function `pnorm()` is similar, but it calculates the cdf instead of the
pdf. Thus, the output consists of actual probabilities rather than probability
densities. Try the following and think about what each one is calculating
(discuss and compare your ideas with your classmates):

```{r}
#| eval: false
pnorm(0)
pnorm(1)
pnorm(1, lower.tail = FALSE)
pnorm(0, mean = 2, sd = 3)
x <- seq(-20, 20, by = 0.1)
y <- pnorm(x)
ggplot() + aes(x = x, y = y) + geom_point()
y <- pnorm(x, mean = 3, sd = 4)
ggplot() + aes(x = x, y = y) + geom_point()
```

The function `qnorm()` does the inverse of `pnorm()`. The input needs to be a
probability, and it will return the quantile that corresponds to that
probability. Try the following and think about what each one is calculating
(discuss and compare your ideas with your classmates):

```{r}
#| eval: false
qnorm(c(0.25, 0.5, 0.75), mean = 1, sd = 2)
x <- seq(0, 1, by = 0.05)
y <- qnorm(x)
ggplot() + aes(x = x, y = y) + geom_point()
y <- qnorm(x, mean = 3, sd = 2)
ggplot() + aes(x = x, y = y) + geom_point()
y <- qnorm(x, mean = 3, sd = 0.1)
ggplot() + aes(x = x, y = y) + geom_point()
```

The function `rnorm()` is a bit different to the others. Rather than
calculating some aspect of a normal distribution, it will randomly generate
realisations from it. The first argument to the function is the number of
realisations you want. Try the following and think about what each one is
calculating (discuss and compare your ideas with your classmates):

```{r}
#| eval: false
rnorm(6)
rnorm(6, mean = 3, sd = 3)
y <- rnorm(250)
ggplot() + aes(x = y) + geom_histogram()
y <- rnorm(250, mean = -2)
ggplot() + aes(x = y) + geom_histogram()
```

## Binomial distributions

R has the same set of four functions for working with a binomial distribution,
similar to the ones shown above for the normal distribution. Run
`help("Binomial")` to see the help page that lists them.

You use them in exactly the same way. However, note that the binomial
distribution is discrete rather than continuous, so it will have a probability
mass function (pmf) rather than a probability density function (pdf).

The function `dbinom()` in fact calculates the pmf, despite the 'd' in its
name. That means the following code calculates a true probability, not a
density:

```{r}
dbinom(3, size = 5, prob = 0.4)
```

This calculates $\Pr(X = 3)$ where $X \sim \mathrm{Bi}(5, 0.4)$.

This is possibly a little confusing if you think of the `d-` prefix as an
abbreviation for 'density', but it does at least maintain a consistent
convention across all of the distributions. Namely, for continuous
distributions the `d-` function gives a pdf, and for discrete distributions it
gives a pmf.

The function `pbinom()` calculates a cdf, similar to the corresponding normal
example above. However, since it is discrete we need to remember that the
probability **includes** the specified value. For example:

```{r}
pbinom(3, size = 5, prob = 0.4)
```

This calculates $\Pr(X \leqslant 3)$ where $X \sim \mathrm{Bi}(5, 0.4)$. If
instead we wanted $\Pr(X < 3)$, then we should actually calculate
$\Pr(X \leqslant 2)$.

## Arbitrary finite discrete distributions

You can use the `sample()` function to simulate from any finite discrete
distribution, by specifying a set of values and associated probabilities.

For example, to select a random day of the week, where the chance of a weekend
is 50% and a weekday is 50%, you could use:

```{r}
days <- c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")
prob <- c(  0.1,   0.1,   0.1,   0.1,   0.1,  0.25,  0.25)
sample(days, size = 1, prob = prob)
```

Note that the default is to sample **without** replacement. If you want
independent and identically distributed (iid) realisations then you need so
specify `replace = TRUE`.

For example, to generate 10 coin flips, you could use:

```{r}
sample(c("H", "T"), size = 10, replace = TRUE)
```

If we do not specify the `prob` argument, then R will make each value equally
likely.

If we do not specify the `size` argument, then R will return as many items as
given in the first argument. If we also do not specify the `replace` argument
(or set it to `FALSE`), then this simply means that R will return a random
permutation (a "random shuffle") of the given values. For example, a random
order of the days of the week can be obtained using:

```{r}
sample(days)
```

## Exercises

1.  Calculate the following probabilities using R.

    a)  The probability of getting exactly 4 heads when flipping a fair coin 8
        times.
    b)  The probability of getting at most 15 heads when flipping a fair coin
        28 times.
    c)  The probability of getting more than 12 heads when flipping a fair coin
        18 times.
    d)  The probability of getting an odd number of heads when flipping a fair
        coin 14 times.

2.  Explore the Poisson distribution. Find its help page. Calculate the
    following for a Poisson distribution with parameter $\lambda = 5$:

    a)  The probability of observing a 3.
    b)  The probability of observing a number of at most 2.
    c)  The probability of observing a number of at least 4.
    d)  Generate 6 realisations.

3.  Calculate the following quantiles using R:

    a)  The 75th percentile (0.75 quantile) of a standard normal.
    b)  The 30th percentile of a t-distribution with 10 degrees of freedom.
    c)  The median of a gamma distribution with shape parameter 2 and rate
        parameter 3.

4.  Let $X_1, X_2, \dots, X_9$ be a random sample. We are told that
    $\mathbb{E}(X_3) = 7$ and $\mathrm{var}(X_4) = 4$.

    a)  What is $\mathrm{sd}(X_2)$?
    b)  What is $\mathrm{var}(X_7 + X_8)$?
    c)  What is an approximate distribution of the sample mean, $\bar{X}$?

5.  Let $X \sim \mathrm{N}(1, 2)$. What is $\mathbb{E}(X^2)$? Approximate this
    by simulating a large number of normal random variables and doing an
    appropriate calculation.

6.  Use simulations to verify the law of large numbers (LLN) when sampling from
    each of the following distributions:

    a)  Normal
    b)  Exponential
    c)  Poisson
    d)  Cauchy

    What do you notice? Can you explain your findings?

# Solving problems via simulation

> In a group of 30 students, what is the probability of that at least two of
> them share the same birthday (month and day)?

This can be tricky to answer correctly using mathematical derivations, but is
trivial to quickly approximate the answer using computer simulations. You saw
this demonstrated in the lectures. Let's try another such problem.

## Lunch for two

> You decide to have lunch at your favourite cafe today, and arrive there at a
> random time between 12pm and 1pm. Your friend also arrives there at a random
> time between 12pm and 1pm. Each of you stay exactly 30 minutes at the cafe.
> What is the probability that the two of you will meet at the cafe?

For convenience, we can we model the arrival time as a number between 0 and 1.
This makes each person's arrival time follow uniform distribution on the
interval $[0,1]$. Let's simulate many such arrival times:

```{r}
N <- 5000
person1 <- runif(N, min = 0, max = 1)
person2 <- runif(N, min = 0, max = 1)
```

The two people will meet if their arrival times are within 30 minutes of each
other, which corresponds to a difference of 0.5 in this mode. Thus, we can
count how often the simulated values differ by 0.5.

```{r}
did_meet <- abs(person1 - person2) < 0.5
mean(did_meet)
```

We see that there's about a 75% chance of meeting.

## Just keep flipping

Sometimes it is easier to use a loop when creating a simulation. Consider the
following problem:

> What's the expected number of coin ﬂips until you get two heads in a row?

We need to simulate coin flips, stopping when we get two in a row, but keeping
a count of how many flips we have done. Let's write a function to carry that
out:

```{r}
flips_simulation <- function() {
  # Initialise counts.
  num_flips <- 0
  num_heads_in_row <- 0
  
  # Keep flipping coins until you get 2 in a row.
  while (num_heads_in_row < 2) {
    
    # Flip a coin.
    new_flip  <- sample(c("H", "T"), size = 1)
    num_flips <- num_flips + 1
    
    # Increment the counts, depending on what you flipped.
    if (new_flip == "H") {
        num_heads_in_row <- num_heads_in_row + 1
      } else {
        num_heads_in_row <- 0
      }
  }

  return(num_flips)
}
```

Now we just need to run that over and over. The handy `replicate()` function
lets us do that easily:

```{r}
num_flips <- replicate(5000, flips_simulation())
mean(num_flips)
```

We expect to flip the coin about 6 times until we get two heads in a row.

## Exercises

7.  You have 20 letters and 20 envelopes. Each letter belongs to a speciﬁc
    envelope. You place the letters randomly in the envelopes, with one letter
    per envelope. What is the expected number of letters that will be in their
    correct envelope?

8.  You take a wooden ruler and break it into two parts at random point
    uniformly along its length. Compare to the original length of the ruler,
    what is the expected length of the smaller part?

9.  You toss a fair coin repeatedly until you get 5 tails in a row. What is the
    expected number of coin tosses?

10. You play the following game: You roll a fair die (equally likely to get any
    of the integers 1 to 6). If the die shows any number from 1 to 5, that is
    your score. However, if it shows a 6, then you score 5 points and get to
    roll the die again. This can continue many times, each time adding the
    number to your score until you no longer roll a 6. What is your expected
    score?

11. Consider again the birthday problem from the lecture. What is the minimum
    number of students you need to have more than a 50% probability of having
    at two students that share a birthday?

# Modelling using probability

The following exercise involves creating a probability model for a given
scenario, and using simulations to explore properties of that scenario.

## Exercise

12. My journey to campus each day involves taking a tram, then a train, then a
    bus. The duration of each leg can be modelled with gamma distributions, as
    follows:

    -   **Tram**: gamma with shape 40 and rate 4
    -   **Train**: gamma with shape 120 and rate 6
    -   **Bus**: gamma with shape 35 and rate 5

    Simulate and visualise the distribution of the total journey time.
    Approximate the mean and standard deviation of this distribution. What's
    the probability that it takes me longer than 40 minutes to get to campus?

# Study design

For the following exercises, you need to use randomisation to complete a task
in the design of each study. Use the `sample()` function in each case.

## Exercises

13. You are investigating the energy efficiency of houses along a major road.
    There are 500 houses along the road, having street numbers consecutively
    from 1 to 500. You won't have time to inspect every house, but can survey
    20 of them to determine their energy ratings. You plan to do this via a
    random sample. Generate a list of 20 houses to inspect.

14. You wish to learn about the experience of students and educators at a
    university, via a survey of 30 people. At this university, there are 5000
    students and 200 educators. In the personnel database, the students have ID
    numbers 1 to 5000, and the educators have ID numbers 9001 to 9200. For your
    survey, you want half of your sample to be students and half to be
    educators. Design a suitable sample. (Note: this will require you to
    generate separate random samples of students and educators. This is known
    as a **stratified random sample**.)

15. You have developed a new soft drink and are wondering whether the colour of
    the can makes a difference to how people perceive the taste and their
    willingness to buy it. You wish to examine this via an experiment. You
    recruit 40 participants and have cans in two different colours, green and
    orange.

    a)  Assign the cans randomly to the participants, so that half receive
        green cans and half receive orange cans.
    b)  Your business partner tries to replicate your experiment, using a
        separate set of participants, but instead of randomly allocating the
        cans he allowed the participants to choose which one they wanted. Which
        version of the experiment is more useful, and why?
